\bf Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-15 manuscript/main.tex
illustrative example

	In modern biomedical studies, it is increasingly common to collect heterogeneous data modalities for each subject. For instance, the Emory University renal study \citep{changBayesianLatentClass2020, jangPrincipalComponentAnalysis2021} records both multiple renogram curves (functional data) and multiple renogram variables (vector data) per kidney, capturing distinct yet related physiological information. Such settings call for statistical methods that can jointly model functional and vector-valued covariates to account for both shared and modality-specific variation linked to clinical outcomes.

@@

One possible approach is to extend smoothed functional PCA \citep{Rice1991} to our optimization problem \eqref{def:maximizer_squared_empirical_cov}, leading to the following formulation:
	\begin{equation*}
	\hat{\xi} 
	= 
	(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\xi}})
	= \arg \max_{\xi \in \mathcal{B}^K \times \mathbb{R}^p,
		\| \xi  \|_\mathcal{H} = 1
	} 
	\biggl\{
	\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)
	- \sum \limits_{j=1}^K \lambda_j 
	\int_0^1 \bigl\{ \hat{\xi}_j^{\prime\prime}(t) \bigr\}^2 dt
	\biggr\}
	,
\end{equation*}
's approach of obtaining smoothed functional PCA components, and find $\xi^*$ that maximizes:
	\begin{equation}
		\widehat{\textnormal{Cov}}^2 \left(\langle \mathbf{W}, \xi \rangle_\mathcal{H}, Y \right) - \sum \limits_{k=1}^K \lambda_k \textnormal{PEN} \left(\psi^{(k)} \right),
		\label{eq: penalized squred variance}
	\end{equation}
	subject to $\xi^{*\top} J^\ast \xi^*=1$. The equation \eqref{eq: penalized squred variance} quantifies the trade-off between fidelity to the data (in this case the sample covariance between the $\rho=\langle \mathbf{W}, \xi \rangle_\mathcal{H}$ and $Y$ in the direction of $\xi$) and roughness as measured by the sum of the individual penalty terms in \eqref{eq: roughness penalty}. The smoothing parameters $\{\lambda_k\}_{k=1}^K$ control the relative importance between the two objectives. Note that the penalty term can be written concisely using matrices as:
		\begin{equation}
		\textnormal{PEN} \left( \hat{\xi}_j \right) 
		=  \int_0^1 
		\bigl\{ \hat{\xi}_j^{''}(t) \bigr\}^2 dt = \mathbf{a}^{(k)T} \ddot{J}^{(k)} \mathbf{a}^{(k)},
		\label{eq: roughness penalty}
	\end{equation}
		where the last term is obtained using the basis expansion $\psi^{(k)}(t)=\mathbf{a}^{(k)T}\mathbf{b}^{(k)}(t)$.

@@

	In this paper, we take an alternative strategy that extends the idea of \citet{Silverman1996}. This approach is known to produce appropriately smoothed functional PCA components under milder conditions than the \citet{Rice1991}'s approach by replacing the usual $L^2$-orthonormality constraint with the one that takes account the roughness of functions via the modified inner product. Specifically, the following proposition provides a generalized Rayleigh quotient problem which modifies the $\mathcal{H}$-orthonormality constraint of the hybrid PLS components to incorporate roughness penalty, and thus whose solution corresponds to a smoothed PLS component maximizing $\widehat{\textnormal{Cov}}^2 \left(\langle \mathbf{W}, \xi \rangle_\mathcal{H}, Y \right)$ over the class of all functions satisfying sufficient smoothness.
	\begin{proposition} \label{Proposition: Regularized generalized rayleigh quotient}
		At each $l$-th iteration of the first step of the PLS algorithm involving the residualized response and predictors $Y_i \equiv Y_i^{[l]}$ and $\mathbf{W}_i \equiv \mathbf{W}_i^{[l]}$, consider the basis function expansions of the functional predictors $X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)$, $i=1, \ldots, n$, and those of the functional parts of the PLS component $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k)$, $k=1, \ldots, K$. Recall from \eqref{def:V_star} that  $V^\ast$ is defined as
			\mathbf{V}^\ast = n^{-2}\begin{bmatrix}
				J \widetilde{C}^\top \mathbf{y}\mathbf{y}^\top \widetilde{C} J & J \widetilde{C}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \mathbf{Z} \\
				\mathbf{Z}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{C} J &   \mathbf{Z}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \mathbf{Z}
			\end{bmatrix} = n^{-2} J^\ast \widetilde{C}^{*\top} \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{C}^{*} J^\ast
			\in \mathbb{R}^{(MK+p) \times (MK+p)}.
		and  $\xi^*=(\mathbf{a}^{\top}, \boldsymbol{\theta}^{\top})^\top \in \mathbb{R}^{MK + p}$. Then, the $l$-th smoothed PLS component, $\widehat{\xi} \equiv \widehat{\xi}_l$ can be obtained in the following way.
		We first compute the basis coefficients by solving the optimization problem
		\begin{align}
			\begin{split}
				&\arg \max \limits_{\xi^* \in \mathbb{R}^{MK+p}} \; \frac{\xi^{*\top} \mathbf{V}^\ast \xi^*}{\xi^{*\top}(J^\ast+\Lambda \ddot{J}^\ast) \xi^*} , \quad \textit{or equivalently}, \\ 
				&\arg \max \limits_{\xi^* \in \mathbb{R}^{MK+p}} \;  \xi^{*\top} \mathbf{V}^\ast \xi^* \quad \textnormal{subject to}  \quad \xi^{*\top} (J^\ast+\Lambda \ddot{J}^\ast) \xi^*=1.
				\label{eq: Regularized generalized rayleigh quotient equation}
			\end{split}
		\end{align}
		We then express the smoothed component in terms of the basis functions as
			\widehat{\xi}[\mathbf{t}] = [\widehat{\psi}(\mathbf{t})^\top, \widehat{\boldsymbol{\theta}}^\top]^\top =  B^\ast[\mathbf{t}]^\top \widehat{\xi}^\ast.   
		The $l$-th PLS score $\widehat{\rho}_{i} \equiv \widehat{\rho}_{il}$ can be obtained as 
		\begin{equation}\label{compute_pls_score}
			\widehat{\rho}_i = \langle \mathbf{W}_i, \widehat{\xi} \rangle_\mathcal{H}
	\end{proposition}
	The proof of Proposition \ref{Proposition: Regularized generalized rayleigh quotient} is provided in Appendix \ref{section:proof:eigenproblem_derivation}.
	Here, $\lambda_k$ $(k=1, \ldots, K)$ is a positive smoothing parameter that controls the trade-off between the goodness of fit and the amount of smoothness in $\psi^{(k)}$. Smaller $\lambda_k$ produces PLS components that fit more closely to the observed data, but selecting too small value may cause overfitting. The estimated PLS component reverts back to the unregularized version of Proposition \ref{proposition:eigen_noregul} when $\lambda_k=0$ for all $k$. On the other hand, larger $\lambda_k$ places more emphasis on smoothing $\psi^{(k)}$. In the limit $\lambda_k \rightarrow \infty$, each of the functions is forced to be of the form $\psi^{(k)}(t)=a+bt$ for some constants $a$ and $b$. In practice, $\{\lambda_k\}_{k=1}^K$ as well as the number of PLS components $L$ to be estimated can be chosen by cross-validation based on a certain predictive performance criterion (e.g., mean squared error).

>>>>

	In modern biomedical studies, it is increasingly common to collect heterogeneous data modalities for each subject. As an illustrative example we use the Emory University renal study \citep{changBayesianLatentClass2020, jangPrincipalComponentAnalysis2021}. The data records both multiple renogram curves (functional data) and multiple renogram variables (vector data) per kidney, capturing distinct yet related physiological information. Such settings call for statistical methods that can jointly model functional and vector-valued covariates to account for both shared and modality-specific variation linked to clinical outcomes.

@@

One possible approach is to extend smoothed functional PCA \citep{Rice1991} to our optimization problem \eqref{def:maximizer_squared_empirical_cov}, leading to the following formulation:
	\begin{equation*}
	\hat{\xi} 
	= 
	(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\xi}})
	= \arg \max_{\xi \in \mathcal{B}^K \times \mathbb{R}^p,
		\| \xi  \|_\mathcal{H} = 1
	} 
	\biggl\{
	\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)
	- \sum \limits_{j=1}^K \lambda_j 
	\int_0^1 \bigl\{ \hat{\xi}_j^{\prime\prime}(t) \bigr\}^2 dt
	\biggr\}
	,
\end{equation*}
where the smoothing parameters $\{\lambda_j\}_{j=1}^K$ control the trade-off between covariance maximization and roughness penalization.
However,   we take an alternative strategy that extends the idea of \citet{Silverman1996}. This approach is known to produce appropriately smoothed functional PCA components under milder conditions than the \citet{Rice1991}'s approach by replacing the usual $L^2$-orthonormality constraint with the one that takes account the roughness of functions via the modified inner product. Specifically, the following proposition provides a generalized Rayleigh quotient problem which modifies the $\mathcal{H}$-orthonormality constraint of the hybrid PLS components to incorporate roughness penalty, and thus whose solution corresponds to a smoothed PLS component maximizing $\widehat{\textnormal{Cov}}^2 \left(\langle \mathbf{W}, \xi \rangle_\mathcal{H}, Y \right)$ over the class of all functions satisfying sufficient smoothness.
	\begin{proposition} \label{Proposition: Regularized generalized rayleigh quotient}
		At each $l$-th iteration of the first step of the PLS algorithm involving the residualized response and predictors $Y_i \equiv Y_i^{[l]}$ and $\mathbf{W}_i \equiv \mathbf{W}_i^{[l]}$, consider the basis function expansions of the functional predictors $X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)$, $i=1, \ldots, n$, and those of the functional parts of the PLS component $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k)$, $k=1, \ldots, K$. Recall from \eqref{def:V_star} that  $V^\ast$ is defined as
		\begin{equation*}
			\mathbf{V}^\ast = n^{-2}\begin{bmatrix}
				J \widetilde{C}^\top \mathbf{y}\mathbf{y}^\top \widetilde{C} J & J \widetilde{C}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \mathbf{Z} \\
				\mathbf{Z}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{C} J &   \mathbf{Z}^\top \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \mathbf{Z}
			\end{bmatrix} = n^{-2} J^\ast \widetilde{C}^{*\top} \widetilde{\mathbf{y}} \widetilde{\mathbf{y}}^\top \widetilde{C}^{*} J^\ast
			\in \mathbb{R}^{(MK+p) \times (MK+p)}.
		\end{equation*}
		and  $\xi^*=(\mathbf{a}^{\top}, \boldsymbol{\theta}^{\top})^\top \in \mathbb{R}^{MK + p}$. Then, the $l$-th smoothed PLS component, $\widehat{\xi} \equiv \widehat{\xi}_l$ can be obtained in the following way.
		We first compute the basis coefficients by solving the optimization problem
		Note that the penalty term can be written concisely using matrices as:
		\begin{equation}
			\textnormal{PEN} \left( \hat{\xi}_j \right) 
			=  \int_0^1 
			\bigl\{ \hat{\xi}_j^{''}(t) \bigr\}^2 dt = \mathbf{a}^{(k)T} \ddot{J}^{(k)} \mathbf{a}^{(k)},
			\label{eq: roughness penalty}
		\end{equation}
		where the last term is obtained using the basis expansion $\psi^{(k)}(t)=\mathbf{a}^{(k)T}\mathbf{b}^{(k)}(t)$.

@@

		\begin{align}
			\begin{split}
				&\arg \max \limits_{\xi^* \in \mathbb{R}^{MK+p}} \; \frac{\xi^{*\top} \mathbf{V}^\ast \xi^*}{\xi^{*\top}(J^\ast+\Lambda \ddot{J}^\ast) \xi^*} , \quad \textit{or equivalently}, \\ 
				&\arg \max \limits_{\xi^* \in \mathbb{R}^{MK+p}} \;  \xi^{*\top} \mathbf{V}^\ast \xi^* \quad \textnormal{subject to}  \quad \xi^{*\top} (J^\ast+\Lambda \ddot{J}^\ast) \xi^*=1.
				\label{eq: Regularized generalized rayleigh quotient equation}
			\end{split}
		\end{align}
		We then express the smoothed component in terms of the basis functions as
		\begin{equation*}
			\widehat{\xi}[\mathbf{t}] = [\widehat{\psi}(\mathbf{t})^\top, \widehat{\boldsymbol{\theta}}^\top]^\top =  B^\ast[\mathbf{t}]^\top \widehat{\xi}^\ast.   
		\end{equation*}
		The $l$-th PLS score $\widehat{\rho}_{i} \equiv \widehat{\rho}_{il}$ can be obtained as 
		\begin{equation}\label{compute_pls_score}
			\widehat{\rho}_i = \langle \mathbf{W}_i, \widehat{\xi} \rangle_\mathcal{H}
		\end{equation}
	\end{proposition}
	The proof of Proposition \ref{Proposition: Regularized generalized rayleigh quotient} is provided in Appendix \ref{section:proof:eigenproblem_derivation}.
	Here, $\lambda_k$ $(k=1, \ldots, K)$ is a positive smoothing parameter that controls the trade-off between the goodness of fit and the amount of smoothness in $\psi^{(k)}$. Smaller $\lambda_k$ produces PLS components that fit more closely to the observed data, but selecting too small value may cause overfitting. The estimated PLS component reverts back to the unregularized version of Proposition \ref{proposition:eigen_noregul} when $\lambda_k=0$ for all $k$. On the other hand, larger $\lambda_k$ places more emphasis on smoothing $\psi^{(k)}$. In the limit $\lambda_k \rightarrow \infty$, each of the functions is forced to be of the form $\psi^{(k)}(t)=a+bt$ for some constants $a$ and $b$. In practice, $\{\lambda_k\}_{k=1}^K$ as well as the number of PLS components $L$ to be estimated can be chosen by cross-validation based on a certain predictive performance criterion (e.g., mean squared error).
